---
title: "How to estimate elastic shape means with elastes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to estimate elastic shape means with elastes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette describes basic usage of elastes in R. 
elastes is a R package that provides functions for the computation of functional 
elastic shape means over sets of open planar curves. The package is particularly 
suitable for settings where these curves are only sparsely and irregularly observed. 
It uses a novel approach for elastic shape mean estimation, where planar curves are
treated as complex functions and a full Procrustes mean is estimated from the
corresponding smoothed Hermitian covariance surface. This is combined with the 
methods for elastic mean estimation proposed in 
[Steyer, Stöcker, Greven (2022)](https://doi.org/10.1111/biom.13706).
See [Stöcker et. al. (2022)](https://doi.org/10.48550/arXiv.2203.10522) for details.


## Installation

Like many other R packages, the simplest way to obtain elastes is to install it directly from CRAN. Type
the following command in your R console:

```{r, eval=FALSE}
install.packages("elastes")
```


## Quick Start

The purpose of this section is to give users a general sense of the package. We will briefly go over the main
functions, basic operations and outputs

Firstly, load up the elastes package.

```{r}
library(elastes)
```

We load a set of 30 sparse handwritten digit 3's from the shapes package for illustration. 
You may have to install the shapes package first.

```{r}
# install.packages("shapes")
digit3.dat <- shapes::digit3.dat
```

The data still has the form of a three dimensional array.
We will have to convert it to a list of data.frames, where each data.frame corresponds to one curve.

```{r}
digit3 <- apply(digit3.dat, MARGIN = 3, FUN = function(curve){
  data.frame(X1 = curve[,1], X2 = curve[,2])
})
```

We can compute a functional mean curve by calling `compute_elastic_shape_mean` 
using the specified set of knots and type of mean (polygonal or smooth).

```{r}
mean_smooth <- compute_elastic_shape_mean(digit3, knots = seq(0, 1, length = 11), type = "smooth")
mean_poly <- compute_elastic_shape_mean(digit3, knots = seq(0, 1, length = 13), type = "polygon")
```

Then we can plot the result. The call to `plot.elastic_shape_mean` shows the estimated
mean function, together with the so called Procrustes fits - the centered and rotation
plus scaling aligned data curves.

```{r, fig.height = 3.5, fig.width = 7, fig.align = "center"}
par(mfrow=c(1, 2))
plot(mean_smooth, main = "smooth mean")
plot(mean_poly, col = "blue", main = "polygonal mean")
```


## Additional parameters

The estimated mean is the leading eigenfunction of the smoothed Hermitian covariance surface of the data curves.
In the covariance estimation, the used penalty can be controlled by the `penalty` parameter.

The method can control for different types of measurement-error variance along t, by setting the `var_type` parameter.

See also the documentation for more information.

```{r, eval=FALSE}
help(compute_elastic_shape_mean)
```
